Rotation Period Evolution in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking

### 2019-02-06 Paper/sync.tex

We examine stellar angular momentum evolution in low-mass binaries by performing a Monte Carlo study of four sets of 10,000 stellar binaries, one modeled using the CPL model and the other using the CTL formalism, using both magnetic braking models.  We simulate both stars' spin evolution but mainly consider the P$_{rot}$ evolution for the primary, i.e. more massive star in binaries, as it is observationally easier to measure a P$_{rot}$ on the more massive, and hence brighter, star \citep[e.g.][]{Meibom2006,Lurie2017}. For each simulation, we sample the primary's mass uniformly over $[0.1, 1]$ M$_{\odot}$. Following \citet{Matt2015}, we uniformly sample the log$_{10}$ of P$_{rot}$ over [$0.8,15$] days, a distribution that approximates the P$_{rot}$ distribution of young stars in the ${\sim}2$ Myr old Orion Nebula Cluster \citep{Stassun1999,Herbst2001,Herbst2002,Rodriguez-Ledesma2009}.  We compute the secondary star's mass by uniformly sampling the mass ratio over [$0.1, 1$] following observations of mass ratios in low-mass binaries \citep{Raghavan2010,Moe2018}. Given the inherent uncertainty in and complexity of the formation of short-period binaries \citep[e.g.][]{Bonnell1994,Bate2000,Bate2002,Moe2018} and the potential for dynamical processing via tides or stellar close encounters \citep[e.g.][]{Mardling2001,Hurley2002,Ivanova2005,Meibom2005}, we take an agnostic approach to the initial orbital configuration by uniformly randomly sampling the initial eccentricity ($e$) over [$0.0,0.3$] and uniformly sampling the initial P$_{orb}$ over [$3,100$] d. Although the CTL model is applicable for $e > 0.3$, the CPL model is not, so we restrict $e \leq 0.3$ to allow us to compare both models. We do not consider P$_{orb} < 3$ d as these binaries are likely to have a tertiary companion \citep{Tokovinin2006} which can significantly impact the inner binary's dynamical evolution \citep[e.g.][]{Munoz2015,Martin2015b,Hamers2016,Moe2018}.

@@

P$_{rot}$ eventually decreases back towards the tidally-locked state in the long-term due to a combination of three simultaneous physical effects.  First, as P$_{rot}$ continue to slow due to magnetic braking, magnetic braking itself weakens as its torque scales as $\omega$ or $\omega^3$ \citep{Kawaler1988}, depending on if the stars are in the saturated or unsaturated regime, respectively \citep{Matt2015}.  Second, as P$_{rot}$ slows down further from the tidally-locked state, tidal torques strengthen as they try to force P$_{rot}$ back towards P$_{eq}$ (see Eqn.~(\ref{eqn:ctl:omega}), for example). Third, when P$_{rot} > $ P$_{eq}$, tides transfer angular momentum from the orbit into stellar rotations, decreasing P$_{orb}$ and strengthening tidal torques that strongly depend on the binary separation as $a^{-6.5}$.  These effects combine to shift the balance of power from magnetic braking-controlled stellar spin down to tidal torques spinning-up stars, shepherding them towards P$_{eq}$ in the long term.

@@

Tidal torques pose a fundamental problem for inferring ages of stars via gyrochronlogy. Regardless of the choice of equilibrium tidal model or magnetic braking model, stellar binaries readily tidally-lock, or at least strongly tidally-interact, across a wide range of P$_{orb}$ and primary star masses, decoupling P$_{rot}$ from age. For example, if one observed a rapidly rotating star with P$_{rot} \lsim 20$ d, gyrochronology models would predict ages $\lsim 2$ Gyr. We highlight this point in Fig.~\ref{fig:ageHist} by plotting a histogram of system ages from Fig.~\ref{fig:protDist} for single or primary stars in binaries with P$_{rot} < 20$ d.  If a rapidly-rotating star is actually an unresolved binary, it would likely be tidally-locked, decoupling P$_{rot}$ from age, causing the predictions of gyrochronology models to fail as its age could be at least several Gyrs. This effect is most likely to manifest in rapid rotators (P$_{rot} < 20$ d), e.g. Fig.~\ref{fig:protAgeHist}, but persists across all P$_{rot}$ up to 100 d, producing a significant contaminating signal, e.g. Fig.~\ref{fig:protDist} and Fig.~\ref{fig:protDistReiners}.

@@

>>>>

We examine stellar angular momentum evolution in low-mass binaries by simulating four sets of 10,000 stellar binaries, one modeled using the CPL model and the other using the CTL formalism, using both magnetic braking models.  We simulate both stars' spin evolution but mainly consider the P$_{rot}$ evolution for the primary, i.e. more massive star in binaries, as it is observationally easier to measure a P$_{rot}$ on the more massive, and hence brighter, star \citep[e.g.][]{Meibom2006,Lurie2017}. For each simulation, we sample the primary's mass uniformly over $[0.1, 1]$ M$_{\odot}$. Following \citet{Matt2015}, we uniformly sample the log$_{10}$ of P$_{rot}$ over [$0.8,15$] days, a distribution that approximates the P$_{rot}$ distribution of young stars in the ${\sim}2$ Myr old Orion Nebula Cluster \citep{Stassun1999,Herbst2001,Herbst2002,Rodriguez-Ledesma2009}.  We compute the secondary star's mass by uniformly sampling the mass ratio over [$0.1, 1$] following observations of mass ratios in low-mass binaries \citep{Raghavan2010,Moe2018}. Given the inherent uncertainty in and complexity of the formation of short-period binaries \citep[e.g.][]{Bonnell1994,Bate2000,Bate2002,Moe2018} and the potential for dynamical processing via tides or stellar close encounters \citep[e.g.][]{Mardling2001,Hurley2002,Ivanova2005,Meibom2005}, we take an agnostic approach to the initial orbital configuration by uniformly randomly sampling the initial eccentricity ($e$) over [$0.0,0.3$], consistent with eccentricities of field binaries that likely have not been tidally-processed \citep{Raghavan2010}. Although the CTL model is applicable for $e > 0.3$, the CPL model is not, so we restrict $e \leq 0.3$ to allow us to compare both models.
We uniformly sample the initial P$_{orb}$ over [$3,100$] d and do not consider P$_{orb} < 3$ d as these binaries are likely to have a tertiary companion \citep{Tokovinin2006} which can significantly impact the inner binary's dynamical evolution \citep[e.g.][]{Munoz2015,Martin2015b,Hamers2016,Moe2018}.

@@

P$_{rot}$ eventually decreases back towards the tidally-locked state in the long-term due to a combination of three simultaneous physical effects.  First, as P$_{rot}$ continue to increase due to magnetic braking, magnetic braking itself weakens as its torque scales as $\omega$ or $\omega^3$ \citep{Kawaler1988}, depending on if the stars are in the saturated or unsaturated regime, respectively \citep{Matt2015}.  Second, as P$_{rot}$ increases further from the tidally-locked state, tidal torques strengthen as they try to force P$_{rot}$ back towards P$_{eq}$ (see Eqn.~(\ref{eqn:ctl:omega}), for example). Third, when P$_{rot} > $ P$_{eq}$, tides transfer angular momentum from the orbit into stellar rotations, decreasing P$_{orb}$ and strengthening tidal torques that strongly depend on the binary separation as $a^{-6.5}$.  These effects combine to shift the balance of power from magnetic braking-controlled stellar spin down to tidal torques spinning-up stars, shepherding them towards P$_{eq}$ in the long term.

@@

Tidal torques pose a fundamental problem for inferring ages of stars via gyrochronlogy. Regardless of the choice of equilibrium tidal model or magnetic braking model, stellar binaries readily tidally-lock, or at least strongly tidally-interact, across a wide range of P$_{orb}$ and primary star masses, decoupling P$_{rot}$ from age. For example, if one observed a rapidly rotating star with P$_{rot} \lsim 20$ d, gyrochronology models would predict ages $\lsim 2$ Gyr. We highlight this point in Fig.~\ref{fig:protAgeHist} by plotting a histogram of system ages from Fig.~\ref{fig:protDist} for single or primary stars in binaries with P$_{rot} < 20$ d.  If a rapidly-rotating star is actually an unresolved binary, it would likely be tidally-locked, decoupling P$_{rot}$ from age, causing the predictions of gyrochronology models to fail as its age could be at least several Gyrs. This effect is most likely to manifest in rapid rotators (P$_{rot} < 20$ d), e.g. Fig.~\ref{fig:protAgeHist}, but persists across all P$_{rot}$ up to 100 d, producing a significant contaminating signal, e.g. Fig.~\ref{fig:protDist} and Fig.~\ref{fig:protDistReiners}.

@@

\subsection{Inferring Binary Tidal Properties from \kepler} \label{sec:inference}

XXX
